Utilizing Parallel Workers: LLVM's Vectorization Plan

### 2025-01-27 paper/paper.tex
paper

\subsection{Vectorization Strategies}
\label{back:vec-strat}
In general, vectorization techniques can be separated into different categories:

@@

\subsection{VPlan Structure}
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

@@

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}.
% TODO: Explain contents of both figures better

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop section of the hierarchical CFG of a VPlan modeling the 
  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
  Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector-trip-count; \texttt{vp<\%2>} = original trip-count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

@@

\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions of the 
underlying scalar IR and will materialize into one or more LLVM IR instructions when executed.

There are many different types of \texttt{VPRecipes}, most notably:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
  % TODO: Explain some instructions more in detail (e.g. CANONICAL-INDUCTION, SCALAR-STEPS)
\end{itemize}

The \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have a \texttt{PREFIX}
in front of them indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related. The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
This block decides if the scalar epilogue loop will be executed following the vectorized loop.

@@

% TODO: VPlan Transformations example

\section{State of VPlan (LLVM 19.1.4)}
\emergencystretch 6em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

To better understand how VPlan is already being used in LLVM, we will now follow along the 
LoopVectorize pass and summarize its key components.

\subsection{Phase 1: Legality}
The main part of this phase is checking whether it is even legal to optimize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

@@

@@

\subsection{Phase 2: Planning}
% TODO: Connect better to previous phase (expand a bit on how the legality artifacts are used when planning)
After the Legality Phase has analyzed the scalar loop, initial VPlans are constructed.
A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

@@

\paragraph{Cost Model}
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally just lookup tables influenced by the underlying 
compile target.
% TODO: Expand on this and explain a bit more

\subsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
First, a vectorized code skeleton is created in the IR, and then the VPlan is executed. 
Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
of its \texttt{VPRecipes} in order.

% TODO: How exactly is the translation from VP recipe back to LLVM IR working? Where does the target
%       backend come in to decide what actual instruction?

Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM and, for branches, 
the general branch condition per block.

@@

If the epilogue loop could be further vectorized, it would be in a second pass by creating VPlans, 
finding the best one, and executing it.

@@

\subsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions.

@@

% TODO: Write about potential future work benchmarking
%       Loop Code Benchmark collection
%       Quick note about collaborative development (example stability & miscompile policy)
%       LLVM Loop Optimization as one of the most active subareas of LLVM development

>>>>

\subsection{Vectorization Strategies}
\label{back:vec-strat}
In general, vectorization strategies can be separated into different categories:

@@

\subsection{VPlan Structure}
\emergencystretch 2em
Before explaining how far this system has been implemented in LLVM, we will review
the different elements of a \texttt{VPlan} in more detail.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=0.475\textwidth]{images/inner-loop-vplan-hcfg-loop-body-color.png}
  \caption{The vectorized loop part of the hierarchical CFG of a VPlan modeling the 
  conditional loop depicted in Figure~\ref{fig:inner-loop-vec}. 
  Surrounding \texttt{VPIRBasicBlock}-nodes are omitted.
  Some predefined values are
  \texttt{vp<\%0>} = VF * UF; \texttt{vp<\%1>} = vector trip count; \texttt{vp<\%2>} = original trip count.}
  \label{fig:inner-loop-vplan-hcfg-body}
\end{figure}

Figure~\ref{fig:inner-loop-scalar-ir} shows the loop part of the scalar IR for the code example
of Figure~\ref{fig:inner-loop-vec}. Comparatively, Figure~\ref{fig:inner-loop-vplan-hcfg-body}
shows the vectorized loop part of the H-CFG of a VPlan that models the vectorization approach described
in Figure~\ref{fig:inner-loop-vec}. Note that the full H-CFG of this VPlan consists of many more nodes 
to handle legality checks (coming before the vectorized loop) or the entire scalar epilogue loop 
(coming after the vectorized loop).

The H-CFG in Figure~\ref{fig:inner-loop-vplan-hcfg-body} consists of 3 nodes on the highest layer. 
\texttt{vector.preheader} and \texttt{middle.block} 
are simple \texttt{VPBasicBlocks} surrounding the main vector loop node. This \texttt{vector loop} node 
is a \texttt{VPRegionBlock} that does 
not replicate its contents (indicated by \texttt{<x1>}; otherwise, it would be \texttt{<xVFxUF>}).

Inside \texttt{vector loop} is only a single \texttt{VPBasicBlock} node \texttt{vector.body}. This block 
models the entire vectorized loop via \texttt{VPRecipes}.

@@

\texttt{VPRecipes} can be seen as instructions of an IR specialized for auto-vectorization. 
A single \texttt{VPRecipe} inside a VPlan is based on zero or more LLVM IR instructions 
of the underlying scalar IR, called ingredients, and will materialize into one or more LLVM IR 
instructions when executed.

There are currently 40 different final types of \texttt{VPRecipes}. The types used 
within Figure~\ref{fig:inner-loop-vplan-hcfg-body} and the most important ones are:
\begin{itemize}
  \item \texttt{VPReplicateRecipe} will replicate one IR instruction either once (\texttt{CLONE})
  or VF * UF times (\texttt{REPLICATE}) into the vectorized IR when executed.
  \item \texttt{VPWidenRecipe} will \texttt{WIDEN} one IR instruction into a
  vector-equivalent when executed.
  \item \texttt{VPInstruction} has no underlying IR instruction it is based on and is used 
  to model newly needed instructions for the vectorized IR. When executed it will 
  \texttt{EMIT} one or more new IR instructions.
  \item \texttt{VPHeaderPHIRecipe} is a sub-type that models any induction, such as first-order 
  recurrence, pointer induction, or reduction as a phi node.
  \item \texttt{VPCanonicalIVPHIRecipe} derives from \texttt{VPHeaderPHIRecipe} and represents a 
  \texttt{CANONICAL-INDUCTION} variable via a \textit{scalar} value.
  \item \texttt{VPScalarIVStepsRecipe} models the individual offsets (\texttt{SCALAR-STEPS}) per 
  vector-lane from a common Integer or Floating-Point induction value.
\end{itemize}

Most of the \texttt{VPRecipes} in Figure~\ref{fig:inner-loop-vplan-hcfg-body} have capitalized 
\texttt{HINTS} indicating their type. Additionally instructions/recipes of the same color 
between Figure~\ref{fig:inner-loop-scalar-ir} and Figure~\ref{fig:inner-loop-vplan-hcfg-body}
are related in their semantical effect.

The only recipes in our example VPlan CFG that do not semantically model any 
instructions found in the scalar IR are from the \texttt{middle.block} (purple). 
This block decides if the scalar epilogue loop is executed after the vectorized loop by 
comparing the vector trip count to the original loop trip count. This additional vector trip count is 
needed as the original loop trip count might not be a perfect multiple of VF * UF, the number of elements 
processed in a single iteration.

The red recipes generally handle the induction variable \texttt{i} and the termination of the 
vectorized loop. \texttt{vp<\%3>} is this loop's scalar induction variable, which goes from 0 to 
\texttt{vp<\%1>}, the vector trip count. The per vector-lane offset's step size is 1 (\texttt{ir<1>}), 
meaning that each 
vector-lane offset from the base induction variable is one greater than the previous lane (we access the 
memory consecutively).

The blue recipes model the condition guarding the addition. \texttt{vp<\%6>} holds the mask, which 
results from this condition and is used for the guarded memory operations.

Finally, the green recipes represent the addition. First, the data from both \texttt{y} and \texttt{x} 
gets loaded (guarded by the compare mask), then added together, and then stored back into 
\texttt{x} (again, guarded by the compare mask).

@@

Such transformations include for example:
\begin{itemize}
  \item Simplifying \texttt{VPRecipes} by removing blend operations with only one unique input, removing 
  redundant type casts, and simplifying logical patterns.
  \item Removing dead \texttt{VPRecipes} whose values have no users.
  \item Moving loop-invariant \texttt{VPRecipes} out of the vector loop region. A \texttt{VPRecipe} is 
  considered loop-invariant if it does not have any side effects, does not read from memory, is not a 
  phi node, and all its operands are defined outside of the vector loop region.
\end{itemize}

\section{State of VPlan (LLVM 19.1.4)}
\emergencystretch 3em
In its current state, VPlan is already deeply integrated into LoopVectorize (LV) and is
used to model, optimize, and execute vectorization of innermost loops~\cite{llvmvplan,llvmintrvplan,llvmvplanupdate}. 
While outer loop vectorization via VPlan has already seen some development, any 
outer loop vectorization features are still experimental and must be explicitly enabled 
by passing the \texttt{-enable-\allowbreak vplan-\allowbreak native-\allowbreak path} 
feature flag~\cite{llvmouterloop,llvmouterloopstatus}.

To better understand how VPlan is already being used in LLVM, we will now follow along the 
\texttt{LoopVectorizePass} and summarize its key components.

\subsection{Phase 1: Legality}
The main part of this first phase is checking whether it is even legal to vectorize a given loop. 
In addition to the general vectorization legality concepts mentioned in 
Section~\ref{sec:vec-legal}, LV has a several extra conditions to check.

@@

Additionally, during this phase, all recurrences of the loop, such as reductions or inductions, are analyzed,
and their phi nodes are stored in a set.

@@

\subsection{Phase 2: Planning}
After the Legality Phase has analyzed the scalar loop and it is deemed legal to vectorize the 
\texttt{LoopVectorizationCostModel} and the \texttt{LoopVectorizationPlanner} are initialized.
The latter one then constructs the initial VPlans based on the properties gathered during the legal phase
and the cost model initialization, such as the recurrence variables or the maximum viable vectorization 
factor.

A VPlan is initialized by creating the H-CFG skeleton building blocks like preheaders, a 
region block for the vector loop, a middle block with the potential runtime check for the scalar 
epilogue loop, and a block for the scalar loop. After the H-CFG has been created, the scalar LLVM IR 
is iterated in topological order, and each scalar IR instruction is transformed into a \texttt{VPRecipe}.
After modeling each scalar IR instruction with a \texttt{VPRecipe}, several VPlan transformations 
are executed, optimizing the initialized VPlan.

@@

\paragraph{Cost Model}
\emergencystretch 4em
Currently, LV and VPlan still mostly rely on the old cost model. This legacy cost model derives costs 
based on the scalar LLVM IR instructions or, if applicable, their vectorized equivalent.
This is suboptimal for VPlan as VPlans work on \texttt{VPRecipes} abstracted away from the underlying 
IR. These costs per instruction are then generally lookup tables influenced by the chosen target 
architecture.

The before-mentioned \texttt{LoopVectorizationCostModel} is a remainder of the old \texttt{LoopVectorize} 
implementation before VPlan. It is a single object that computes different costs based on the scalar loop 
IR without any relation to different VPlans. For example, it stores the instructions that will remain scalar 
after vectorization or the LLVM values for which the cost can 
be ignored. The latter includes, for instance, loop invariant operations that would be hoisted out of the loop 
or trivially dead instructions.

Additionally, each \texttt{VPRecipe} has a function to calculate its own cost. For most \texttt{VPRecipes} 
that reference one or more LLVM IR ingredients, these are then used to calculate the cost of the recipe.
On the other hand, \texttt{VPRecipes} that do not directly model an underlying LLVM scalar IR instruction, 
such as \texttt{VPInstruction}, can generally not compute their own cost and thus return a cost of zero 
for now.

For simple innermost loops with few vectorization approaches,
relying on the old cost model might not have too much of a performance impact. 
However, once more VPlan transformations and complex
outer/hybrid loop vectorization are implemented, this new cost model would be an important
detail of the VPlan system to reliably choose the most optimal vectorization approach.

\subsection{Phase 3: Execution}
After the best VPlan has been chosen, it is executed (\texttt{LoopVectorizationPlanner::executePlan}).
For this VPlan, some final transformations are run depending on a specifically chosen VF and UF.
Then a vectorized code skeleton is created in the IR, and the VPlan executed. 
Upon execution, the H-CFG of the VPlan is walked, and each node is executed, which in turn executes all 
of its \texttt{VPRecipes} in order. Each \texttt{VPRecipe} produces its respective output IR.

A \texttt{VPTransformState} object is passed along to store information 
needed for generating the output IR, such as the chosen vectorization factor or generated values that were 
used by later \texttt{VPRecipes}.

@@

While Figure~\ref{fig:inner-loop-vec-cfg} shows the general CFG of a vectorized loop in LLVM, some 
nodes can also be omitted in special cases.

For example, if all the pointers used inside the function are marked as \texttt{noalias}, the compiler 
is guaranteed that they will not overlap and the \texttt{vector.memcheck} block will be omitted.
While the C language and most modern C++ compilers have an explicit \texttt{restrict} keyword for this,
Rust automatically applies the \texttt{noalias} attribute to safe references when compiling via LLVM.

Additionally, before the initial VPlans are constructed at the beginning of Phase 2, it is also checked if 
the scalar epilogue loop could be folded into the vectorized loop version itself by increasing the vector 
trip count over the actual loop trip count and masking all operations. This tail-folding approach in LLVM 
has not yet been enabled by default.

Otherwise, if the scalar epilogue loop could be vectorized, it would be in a second loop-vectorize 
pass by, again, creating VPlans, finding the best one, and executing it.

@@

\subsection{VPlan and Integration with LV}
Setting up and integrating the VPlan structure into LV has seen tremendous effort over the last 
few years. Florian Hahn presented future directions for 
VPlan~\cite{llvmvplanupdate}, which mostly revolve around transforming more LV-based decisions into
explicit VPlan-to-VPlan transformations. Additionally, the legacy LV cost model still influences 
many cost model-related decisions. However, the transition to a VPlan-based cost model has already 
started. This new model would derive costs based on \texttt{VPRecipes} directly instead of the 
abstracted scalar IR instructions. Besides that, the community also focuses on enabling tail folding by
default, which would lead to a significant performance increase in some instances.

@@

% TODO: Write about potential future work benchmarking
%       Loop Code Benchmark collection
%       Quick note about collaborative development (example stability & miscompile policy)
%       LLVM Loop Optimization as one of the most active subareas of LLVM development
% TODO: "Also, it would be great to have the main problem and paper’s aim to be re-stated in the summary section for the first reading pass."
%       "summary lacks the problem statement and the big picture"
